Lead Contamination Testing: Is Water in Toronto Safe for Drinking?

### 2024-09-27 paper/paper.qmd
final formatting of the paper

## Data Set and Context
The data set "Non Regulated Lead Sample" by Open Data Toronto aims to
address the concerns Toronto residents have for their drinking water.
The data set contains information on lead concentrations in water
samples. Each row represents a water sample, with corresponding information
on the date, partial postal code where the sample was provided, and the
main variable of interest, lead concentration in parts per million
(ppm).

@@

**Date**: Represents the date on which the water sample was collected,
can be used to detect trends and seasonal variations. It is evident from (@fig-showdate) that the data points are evenly distributed through time except during 2020, there is an empty period of time. This might be caused by COVID-19 quarantine, where the residence does not have the chance to volunteer their water sample.
```{r}
#| label: fig-showdate
#| fig-cap: "Graph of Dates Categorized by Year and Presented by Each Months"
#| echo: false
# Create a new data frame with extracted year and month
data_date <- data %>%
  transmute(
    Year = year(`Sample Date`),
    Month = month(`Sample Date`, label = TRUE, abbr = TRUE)
  )

@@

**Partial Postal Code:** A truncated form of the postal code indicating
the general area where the water sample was taken, can be used to
identify potential geographical clusters. The (@fig-splitpostal) shows the partial postal code categorized by the first two characters and the y axis being the last character ranging from A-Z. It can be oberseved from (@fig-splitpostal) that the data points are not evenly distributed. This can be caused by sampling bias since it is a volunterring sampling but it can also be caused by the fact that some areas in City of Toronto are not present for residency.
```{r}
#| label: fig-splitpostal
#| fig-cap: "Graph of Partial Postal Code Splitting into First Two Characters and Last Character"
#| echo: false

@@

# Plot the extracted components using ggplot2
ggplot(data_postal, aes(x = FirstTwoChars, y = LastChar)) +
  geom_jitter(width = 0.2, height = 0.2, color = blueishgrey, size = 3) + # Jitter points to spread them out for better visualization
  labs(title = "Relationship Between First Two and Last Characters of Postal Codes",
       x = "First Two Characters",
       y = "Last Character") +
  theme_minimal()

@@

**Lead Concentration (ppm):** The measured amount of lead in the water
sample, expressed in parts per million(@fig-boxplot). This is the primary variable of
interest, as it directly relates to public health risks associated with
lead exposure.Additionally,     combinations of lead concentration with other variables can be furthur analyzed.
```{r}
#| label: fig-boxplot
#| fig-cap: "Boxplot showing the lead concentration above the saftey threshold"
#| echo: false

@@

# Create the boxplot using ggplot2
ggplot(data, aes(y = `Lead Amount (ppm)`)) +
  geom_boxplot(fill = "skyblue") +
  geom_hline(yintercept = threshold_value, color = "red", linetype = "dashed") +
  labs(title = "Boxplot of Lead Amount (ppm) with Threshold Line",
       y = "Lead Amount (ppm)") +
  theme_minimal()

@@

For the purpose of data analysis the raw data set went through some
pre-processing: NA Values: All missing (NA) values were removed to
ensure accurate analysis. Lead Concentration: The Lead Amount (ppm)
variable was cleaned to remove any anomalies and ensure that values were
properly numeric. Date Variable Type: The Date variable was converted to
Date type in R to facilitate temporal analysis and aggregation by months
and years.

@@

## Summary
A summary table(@tbl-summary) for num type variable lead amount.It is evident that the minimum and maximum value differs by 168,800 times, which is very significant. In (@gif-boxplot), it is evident that there are many outliers. Therefore, the assumption can be made that there are water samples that are very clean in terms of lead concentration and there are samples that are contaminated. It is clear that the mean and median differs a lot. Which means that the distribution of this data set is highly right skewed. The outliers will elevated mean lead concentration level in this data set.

@@

```
## Relationships
These relationships are upon further discussion in @sec-mydiscussion.

@@

## Measurement{#sec-measurement}
In this dataset, the aim is to explore in the concentration of lead in drinking water across different regions of a city and over time. The measurement process transforms a real-world phenomenon—lead contamination in water—into quantitative entries in a dataset that can be analyzed for trends.

@@

## Lead Concentration influencing Health{#sec-health}

### WHO and Canadian federal guidance on lead concentration in water

@@

### Percentage of data above this threshold

From calculations by R, there are 97.26% of the data collected indicating that
the lead amount in the water is safe. As shown in (@tbl-threshold), there are only 280 counts of water sample collected that is above MAC, which takes up only 2.74% of the whole data set. Even though this seems like a good result, those 280 measurements still effects people and their family's health since those lead concentrations will have negative effect on health(@w)(@c-1).
```{r, echo=FALSE}
#| label: tbl-threshold
#| tbl-cap: "Number and Percentage of data above Maximum Acceptable Concentration"
#| echo: false
# Define the threshold
threshold <- 0.010

@@

### plot of the data above this threshold
To visualize those data above MAC, data below MAC and above 2 were filtered out, because from (@fig-boxplot), it is clear that there are only 4 data points above 2, therefore, with that in mind it is safe to temporarily ignore to get a scale that will better show the most data points above MAC. A scatterplot was used with x axis being date and y axis being lead concentration give a graph that spreads out lead concentrations instead of on a 1-dimensional line. From (@fig-between) this indicates that there are clusters of data points below 0.25 and for higher lead concentrations, there are less data points and more spread out than those below 0.25.

@@

## How districts affect Lead Concentration{#sec-district}

### graphs of postal code vs lead concentrations

Nine boxplot (@fig-boxplotdistrict) each representing a district, which is the constructed variable from the first two letters of partial postal code. Again, most data points are around 0, and the outliers streched the scale so that this way we can easily analyze the outliers. The districts containing outliers were also shown in (@fig-boxplotdistrict). It is clear that the "M4" has two data points above 8 ppm, as
well as a cluster between 0 and 1 ppm. "M6" also have many data between 0 and
2ppm. It is safe to assume that these two districts have water sources with high lead concentrations.
```{r}
#| label: fig-boxplotdistrict
#| fig-cap: "Boxplot Showing Outliers Categorized by District"
#| echo: false

@@

### extract outliers above the threshold

@@

### mean value of lead concentration for different districts

@@

```
\newpage
```{r}
#| label: fig-meandistrict
#| fig-cap: "Mean Lead Concentration Categorized by First Two Characters from Partial Postal Code"
#| echo: false
# Step 4: Construct a bar graph using ggplot2
ggplot(mean_lead_by_district, aes(x = District, y = `Lead Amount (ppm)`)) +
  geom_bar(stat = "identity", fill = blueishgrey) +
  labs(title = "Mean Lead Concentration by District",
       x = "District (First Two Characters of Postal Code)",
       y = "Mean Lead Amount (ppm)") +
  theme_minimal()
```

## How months and year affect Lead Concentration{#sec-time}

### Times Series analysis for autocorrelation{#sec-timeseries}

@@

### graphs of months vs mean lead concentration
With all data points, there are no obvious trends over time. The examination is made closely for the constructed variable month from Date. (@fig-month) is the boxplot showing the outliers, but there are not enough evidence to show that there are specific month that will have a higher lead concentration. Therefore, a similar filtering is applied to data larger than 2 ppm.
```{r}
#| label: fig-month
#| fig-cap: "Boxplots that Shows the Outliers Categorized by Months"
#| echo: false
# Extract month as a factor for visualization
data_by_month <- data %>%
  mutate(Month = month(`Sample Date`, label = TRUE, abbr = TRUE))

@@

### graphs of years vs mean lead concentration
Similar to months,variable called year was also constructed. However, both the boxplots and the mean values without data below 2 ppm does not show trend from year to year. This aligns with our conclusion in @sec-timeseries.
```{r}
#| label: fig-year
#| fig-cap: "Boxplots that Shows the Outliers Categorized by Year"
#| echo: false
# Create a new data frame with the extracted year
data_by_year <- data.frame(
  Year = year(data$`Sample Date`),
  lead = data$`Lead Amount (ppm)`
)

@@

```
```{r}
#| label: fig-meanyear
#| fig-cap: "Mean Lead Concentration Without Data Greater than 2 ppm categorized by Year"
#| echo: false
# Filter data with threshold of 2 ppm
filtered_data <- data %>%
  filter(`Lead Amount (ppm)` <= 2)

@@

## Weaknesses and next steps{#sec-future}

A key weakness of the current dataset is that it originates from a variety of water sources and is not necessarily limited to direct drinking water. Since the water sample collection process was not controlled, it is uncertain whether all samples reflect tap water quality. Despite this, the guidelines used for analysis are intended for drinking water safety, which introduces a potential mismatch between the data and its intended use. The program is non-regulated, leaving the sampling process up to residents and without any standardization; as a result, the reliability and comparability of the data across different samples may be compromised.

Considering these weaknesses, it is important to recognize the role of pre-treatment, such as water filtration systems, in affecting lead concentration. This necessitates the documentation of whether samples are taken before or after filtration, which could significantly influence the interpretation of lead levels in relation to health guidelines. To improve the reliability of future data collection, a probability-based adaptive sampling plan should be considered. This approach would ensure more representative sampling of tap water from various districts, providing a better understanding of the exposure to lead in drinking water. Additionally, further analysis on the seasonality of lead levels could reveal patterns influenced by environmental factors such as temperature, water flow rates, and seasonal water treatment processes. Identifying these trends is crucial for developing targeted interventions to reduce lead exposure throughout the year.

@@

# Appendix {.unnumbered}
The color used in graphs :
blueishgrey <- rgb(96, 139, 182, maxColorValue = 255) is the color of lead in reality.

@@

>>>>

## Data Set and Context

The data set "Non Regulated Lead Sample" by Open Data Toronto aims to
address the concerns Toronto residents have for their drinking water.
The data set contains information on lead concentrations in water
samples. Each row represents a water sample, with corresponding
information on the date, partial postal code where the sample was
provided, and the main variable of interest, lead concentration in parts
per million (ppm).

@@

**Date**: Represents the date on which the water sample was collected,
can be used to detect trends and seasonal variations. It is evident from
(@fig-showdate) that the data points are evenly distributed through time
except during 2020, there is an empty period of time. This might be
caused by COVID-19 quarantine, where the residence does not have the
chance to volunteer their water sample.

```{r}
#| label: fig-showdate
#| fig-cap: "Graph of Dates Categorized by Year and Presented by Each Months"
#| echo: false
# Create a new data frame with extracted year and month
data_date <- data %>%
  transmute(
    Year = year(`Sample Date`),
    Month = month(`Sample Date`, label = TRUE, abbr = TRUE)
  )

@@

**Partial Postal Code:** A truncated form of the postal code indicating
the general area where the water sample was taken, can be used to
identify potential geographical clusters. The (@fig-splitpostal) shows
the partial postal code categorized by the first two characters and the
y axis being the last character ranging from A-Z. It can be oberseved
from (@fig-splitpostal) that the data points are not evenly distributed.
This can be caused by sampling bias since it is a volunterring sampling
but it can also be caused by the fact that some areas in City of Toronto
are not present for residency.

```{r}
#| label: fig-splitpostal
#| fig-cap: "Graph of Partial Postal Code Splitting into First Two Characters and Last Character"
#| echo: false

@@

# Plot the extracted components using ggplot2
ggplot(data_postal, aes(x = FirstTwoChars, y = LastChar)) +
  geom_jitter(width = 0.2, height = 0.2, color = blueishgrey, size = 3) + # Jitter points to spread them out for better visualization
  labs(title = "Relationship Between Splitted Postal Codes",
       x = "First Two Characters",
       y = "Last Character") +
  theme_minimal()

@@

**Lead Concentration (ppm):** The measured amount of lead in the water
sample, expressed in parts per million(@fig-boxplot). This is the
primary variable of interest, as it directly relates to public health
risks associated with lead exposure.Additionally, combinations of lead
concentration with other variables can be furthur analyzed.

```{r}
#| label: fig-boxplot
#| fig-cap: "Boxplot showing the lead concentration above the saftey threshold"
#| echo: false

@@

# Create the boxplot using ggplot2
ggplot(data, aes(y = `Lead Amount (ppm)`)) +
  geom_boxplot(fill = "skyblue",outlier.colour = "red") +
  geom_hline(yintercept = threshold_value, color = "red", linetype = "dashed") +
  labs(title = "Boxplot of Lead Amount (ppm) with Threshold Line",
       y = "Lead Amount (ppm)") +
  theme_minimal()

@@

For the purpose of data analysis the raw data set went through some
pre-processing:

**NA Values:** All missing (NA) values were removed to ensure accurate
analysis.

**Lead Concentration:** The Lead Amount (ppm) variable was cleaned to
remove any anomalies and ensure that values were properly numeric.

**Date Variable Type:** The Date variable was converted to Date type in
R to facilitate temporal analysis and aggregation by months and years.

@@

## Summary

A summary table(@tbl-summary) for num type variable lead amount.It is
evident that the minimum and maximum value differs by 168,800 times,
which is very significant. In (@fig-boxplot), it is evident that there
are many outliers. Therefore, the assumption can be made that there are
water samples that are very clean in terms of lead concentration and
there are samples that are contaminated. It is clear that the mean and
median differs a lot. Which means that the distribution of this data set
is highly right skewed. The outliers will elevated mean lead
concentration level in this data set.

@@

## Relationships

These relationships are upon further discussion in @sec-mydiscussion.

@@

## Measurement {#sec-measurement}

In this dataset, the aim is to explore in the concentration of lead in
drinking water across different regions of a city and over time. The
measurement process transforms a real-world phenomenon—lead
contamination in water—into quantitative entries in a dataset that can
be analyzed for trends.

@@

## Lead Concentration Influencing Health {#sec-health}

### WHO and Canadian Federal Guidance on Lead Concentration in Water

@@

### Percentage of Data Above the Threshold

From calculations by R, there are 97.26% of the data collected
indicating that the lead amount in the water is safe. As shown in
(@tbl-threshold), there are only 280 counts of water sample collected
that is above MAC, which takes up only 2.74% of the whole data set. Even
though this seems like a good result, those 280 measurements still
effects people and their family's health since those lead concentrations
will have negative effect on health(@w)(@c-1).

```{r, echo=FALSE}
#| label: tbl-threshold
#| tbl-cap: "Number and Percentage of data above Maximum Acceptable Concentration"
#| echo: false
# Define the threshold
threshold <- 0.010

@@

### Plot of the Data Above the Threshold

To visualize those data above MAC, data below MAC and above 2 were
filtered out, because from (@fig-boxplot), it is clear that there are
only 4 data points above 2, therefore, with that in mind it is safe to
temporarily ignore to get a scale that will better show the most data
points above MAC. A scatterplot was used with x axis being date and y
axis being lead concentration give a graph that spreads out lead
concentrations instead of on a 1-dimensional line. From (@fig-between)
this indicates that there are clusters of data points below 0.25 and for
higher lead concentrations, there are less data points and more spread
out than those below 0.25.

@@

## How Districts Sffect Lead Concentration {#sec-district}

### Graphs of Postal Code vs Lead Concentrations

Nine boxplot (@fig-boxplotdistrict) each representing a district, which
is the constructed variable from the first two letters of partial postal
code. Again, most data points are around 0, and the outliers streched
the scale so that this way we can easily analyze the outliers. The
districts containing outliers were also shown in (@fig-boxplotdistrict).
It is clear that the "M4" has two data points above 8 ppm, as well as a
cluster between 0 and 1 ppm. "M6" also have many data between 0 and
2ppm. It is safe to assume that these two districts have water sources
with high lead concentrations.

```{r}
#| label: fig-boxplotdistrict
#| fig-cap: "Boxplot Showing Outliers Categorized by District"
#| echo: false

@@

### Extract Outliers Above the Threshold

@@

### Mean Values of Lead Concentration for Different Districts

@@

```{r}
#| label: fig-meandistrict
#| fig-cap: "Mean Lead Concentration Categorized by First Two Characters from Partial Postal Code"
#| echo: false
# Step 4: Construct a bar graph using ggplot2
ggplot(mean_lead_by_district, aes(x = District, y = `Lead Amount (ppm)`)) +
  geom_bar(stat = "identity", fill = blueishgrey) +
  labs(title = "Mean Lead Concentration by District",
       x = "District (First Two Characters of Postal Code)",
       y = "Mean Lead Amount (ppm)") +
  theme_minimal()
```

## How Months and Years Affect Lead Concentration {#sec-time}

### Times Series Analysis for Autocorrelation {#sec-timeseries}

@@

### Graphs of Months vs Mean Lead Concentration

With all data points, there are no obvious trends over time. The
examination is made closely for the constructed variable month from
Date. (@fig-month) is the boxplot showing the outliers, but there are
not enough evidence to show that there are specific month that will have
a higher lead concentration. Therefore, a similar filtering is applied
to data larger than 2 ppm.

```{r}
#| label: fig-month
#| fig-cap: "Boxplots that Shows the Outliers Categorized by Months"
#| echo: false
# Extract month as a factor for visualization
data_by_month <- data %>%
  mutate(Month = month(`Sample Date`, label = TRUE, abbr = TRUE))

@@

### Graphs of Years vs Mean Lead Concentration

Similar to months,variable called year was also constructed. However,
both the boxplots and the mean values without data below 2 ppm does not
show trend from year to year. This aligns with our conclusion in
@sec-timeseries.

```{r}
#| label: fig-year
#| fig-cap: "Boxplots that Shows the Outliers Categorized by Year"
#| echo: false
# Create a new data frame with the extracted year
data_by_year <- data.frame(
  Year = year(data$`Sample Date`),
  lead = data$`Lead Amount (ppm)`
)

@@

```{r}
#| label: fig-meanyear
#| fig-cap: "Mean Lead Concentration Without Data Greater than 2 ppm categorized by Year"
#| echo: false
# Filter data with threshold of 2 ppm
filtered_data <- data %>%
  filter(`Lead Amount (ppm)` <= 2)

@@

## Weaknesses and Next Steps {#sec-future}

A key weakness of the current dataset is that it originates from a
variety of water sources and is not necessarily limited to direct
drinking water. Since the water sample collection process was not
controlled, it is uncertain whether all samples reflect tap water
quality. Despite this, the guidelines used for analysis are intended for
drinking water safety, which introduces a potential mismatch between the
data and our threshold maximum acceptable concentration. The program is
non-regulated, leaving the sampling process up to residents and without
any standardization. As a result, the reliability and comparability of
the data across different samples may be varied.

It is important to recognize the role of pre-treatment, such as water
filtration systems, in affecting lead concentration. The result of this
paper might be altered if we can ensure every water sample was
pre-treated to filter out the hard metals.

To improve the reliability of future data analysis, a probability-based
adaptive sampling plan should be considered. This approach would ensure
more representative sampling of tap water from various districts,
providing a better understanding of the exposure to lead in drinking
water. There should also be probability based data analysis to better
estimate correlations between lead concentrations and other variables,
not limited to those mentioned in this paper.

Additionally, further analysis on the seasonality of lead levels could
verify whether there are patterns influenced by environmental factors
such as temperature, water flow rates, and seasonal water treatment
processes. Identifying these trends is crucial for developing targeted
interventions to reduce lead exposure throughout for public health.

@@

# Appendix {.unnumbered}

The color used in graphs : blueishgrey \<- rgb(96, 139, 182,
maxColorValue = 255) is the color of lead in reality.

All outliers were represented by color = "red".

This paper serves the purpose of telling story with data, there is only
minimal analysis performed, for a deeper studies, consider use the
processed data for training and analyzing.

@@

The data can be found in OpenDataToronto with
<https://open.toronto.ca/dataset/non-regulated-lead-sample/>.

The cleaning process includes removing NA, and modifying types of
variables. Tests were performed to ensure that the data is suitable for
analyze.
